Type Checker Generation using Reference Attribute Grammars

### 2024-01-07 MSc.tex
Address many of the comments left by supervisor

\lstdefinestyle{mystyle}{
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}
\lstset{style=mystyle}
\lstdefinelanguage{jrag}[]{java}{
  morekeywords={aspect, syn, eq, inh}
}

@@

\newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}

@@

% For questions to people reviewing, remove before finalizing
\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
%\newcommand{\reviewquestion}[1]{}

@@

\makefrontmatter
\chapter{Introduction} % should give the background of your work. Important parts to cover:
% Give the context of your work, have a short introduction to the area.
While writing software, developers often make mistakes.
Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
\CR{Possible papers to cite as evidence:
  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
  compelling to motivate your work.}
Type checkers are one of the most common static checking methods, ensuring that the correct data types are used, but creating a new type checker is often an arduous process, requiring large amounts of complex code.
Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.

% Define the problem you are solving (or trying to solve).
\CR{Prior work has already looked into pluggable type checkers.
  Why is translating them into JastAdd interesting?
  (aside from theoretical questions, you can also use
  pragmatic reasons, such as integration into our courses.)}
In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of reference attribute grammars.

\CR{Make clear to the reader what the challenges are:
   How are these two formalisms similar, how do they differ (at a high level)?
   You can use a simple typing rule as an example if it helps (very optional).
   Hint: look at determinism in evaluation (syntax-directedness), unification,
   but stick to intuition in this chapter; you can elaborate in Chapter 2 or later
   if you think it helps.}

%TODO: Give an overview of the other sections
\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning
  chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
  We evaluate the approach against Foo, and demonstrate Bar.
  Our case studies show Quux.
  Overall, this initial exploration shows \ldots''}

\CR{paper $\to$ thesis; also, this always comes at the very end of the introduction}
The rest of this paper is outlined as follows.
% This chapter introduces the
Chapter 2 presents the necessary background information for the report.
Chapter 3 describes the design and implementation of our program, followed by demonstrations of its results in chapter 4.
Chapter 5 discusses the results in a more general sense, before the conclusion in chapter 6.

% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
\section{Research questions}
To formulate our goals more concretely, we have \CR{Avoid contractions in technical writing} set out to answer the following research questions:
\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
\newcommand{\rqone}{What kinds of typing rules can we translate through \CR{``through?''} RAG?}
\newcommand{\rqtwo}{How can we implement the translations and into what form?}
\newcommand{\rqthree}{What are the potential issues of translating typing rules into JastAdd-based RAGs?}
\begin{itemize}
  \item \rqone \CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
  \item \rqtwo \CR{What does ``form'' mean, here?  How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
  \item \rqthree \CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
\end{itemize}

\section{Related work} % Maybe move to discussion?
% Describe related work (literature study). Besides listing other work in the area, mention how is it related or relevant to your work. The tradition in some research area is to place this part at the end of the report (check with your supervisor).
Pacak et al. implemented a compiler for generating incremental type checkers using Datalog\cite{Pacak}.
While the goal is similar to ours, they take a different approach, by converting typing rules into the declarative logic programming language \CR{second half of this sentence confuses me}.

Their work is thereby largely focused on expressing typing rules within the limits of the Datalog language, which can only compute finite relations.
We have a significantly different compilation target, JastAdd, which allows us to use reference attribute grammars to evaluate the typing relations, which we write in a more iterative \CR{unclear?} approach.
\CR{Datalog is fully declarative, whereas JastAdd uses declarative rules that connect snippets of imperative Java code.  Is that what you meant?}
\CR{Some more thoughts below to help make the distinction clearer (all optional):}
\CR{Datalog has no direct notion of ASTs or other tree structures / algebraic terms, so trees must be encoded into relations.}
\CR{Datalog has a simple, built-in notion of unification that can handle cases like testing sibling relations: $\textsf{Sibling}(x, y) :- Parent(p, x), \textsf{Parent}(p, y), x \ne y.$.
    This is something you deal with later in JastAdd.}

Additionally, their project defines a formalism for expressing typing rules within the Datalog language, ours defines a language which aims to keep as close as possible to the traditional inference rules used for typing rules, to minimise the difficulty of translation.
\CR{I'd start with this difference-- the translation is targeting two rather different formalisms, which is all you need to justify the novelty.}

\chapter{Background}
%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinkingr ``I know why I want to read this''}
This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
The first section explains abstract syntax trees, which are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
    analysing programs.}
The next section details type systems \CR{Too much detail in the following:}, by first introducing data types, then explaining the components required for a type system, followed by a description of static type checkers.
\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}

\section{Abstract syntax trees}
\CR{What's a syntax?}
An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It's \CR{contraction} abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this information \CR{what is ``this information'', concretely?}.
A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

This layer of abstraction makes it easier to write analyses for the program, as we can focus on the key elements and structure.
\CR{Is there a connection between the AST and the behaviour of the program?}

\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}

\section{Type theory}
In this section we will discuss the elements of type theory required to understand this thesis.
Firstly, we will discuss what types are and give some examples of their use.
Next, we will describe type systems, which form a relation between types and expressions.

\subsection{Data types}
\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
Computer software generally needs to handle and manipulate several different types of data in one program, such as numbers or text.
Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.

\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}

\subsection{Type checking}
There are two main approaches to ensuring that the aforementioned constraints are upheld, dynamic and static type-checking.
\CR{Why would we even want to ensure that they are upheld?}
Dynamic checking is done performed runtime, with each type checked only when needed.
Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}

As our project generates a static type checker, we will focus our explanations on that topic.
Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.
Conservative \CR{incomplete sentence?}

\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
  For each concrete (and possibly very complex) operator on the left hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractoins
  over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
}

\subsection{Type systems}
A type system consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
\CR{Why do the types need a syntax?
    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}

\subsubsection{Type syntax}
A type syntax is a definition of the valid types within a language.
\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
\CR{A bit vague; consider giving an example.}

\subsubsection{Typing rules}
The typing relation, notated \lstinline{t : T}, is a link between the expressions of a language and the types of the type syntax.
\CR{It is \emph{the} link, even.}
The left-hand side is an expression in the object lanuage, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.

Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
The typing rule \lstinline{T-True} seen in figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.

\begin{figure}[h]
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  \caption{Typing rule for a true expression}
  \label{typingruleexampletrue}
\end{figure}

Typing rules may also have premises, a number of conditions which \CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')} must hold for the typing rule to apply.
The notational convention here is to put the premises above a bar, with the conclusion underneath it.

\begin{figure}[h]
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \text{T}\quad
      t_3 : \text{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \text{T}
    }
  \end{equation}
  \caption{Typing rule for an if-expression}
  \label{typingruleexampleif}
\end{figure}

Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
The bottom line, the conclusion, contains the keywords of the if-expression and three variables $t_i$ which represent any term within the language.
Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable T, which represents any single arbitrary type within the type syntax.
While T may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
This T also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
\newcommand{\tyvar}[1]{\ty{\textcolor{cyan!50!black}{#1}}}
\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.
\CR{Please explain the purpose of typing rules in static type checking.  Using just your two examples you would be able to
  show a well-typed program, and if you add one more rule (e.g. for integers), you can show a type error.  No need to build
  the full derivation tree, but it's important to mention recursion here.  This entire bit may be worth a separate subsubsection.}

\section{Reference Attribute Grammars}
This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compiliation system\cite{jastadd}.

\subsection{Reference Attribute Grammars}
Reference Attribute Grammars (RAGs) are an extension to Attribute Grammars (AGs), a formalism used for specifying the semantics of programming languages.
\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.

@@

\subsection{JastAdd}
JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.
\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}

\cite{JastAdd}
\subsubsection{AST specification}\label{astdef}
One of the key components of a JastAdd projects are the \verb|.ast| files, which declare the structure of the abstract syntax tree.
This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
These AST classes can have a variety of forms, supporting both abstract classes and subclasses, and can contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.

\subsubsection{RAG Modules}
\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
not a necessary feature for RAGs.  Please introduce them in a seprate section: you can very directly connect them to the idea of ``type syntax''.}
\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{boolstrgen}
\end{figure}
The RAG modules written in \verb|.jrag| files, declare the attributes to be included in the classes generated by the AST specification.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

@@

%\subsection{Typing rule compiler}\label{typingrulecompiler}

@@

\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {

@@

\section{Research questions}

>>>>

% disable spellcheck for this section
% LTeX: SETTINGS enabled=false
\lstdefinestyle{mystyle}{
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}
\lstset{style=mystyle}
\lstdefinelanguage{jrag}[]{java}{
  morekeywords={aspect, syn, eq, inh}
}
% LTeX: SETTINGS enabled=true

@@

\newcommand{\CR}[1]{\textcolor{green!60!black}{[\textbf{CR}:#1]}}
%\newcommand{\CR}[1]{}

% For questions to people reviewing, remove before finalizing
\newcommand{\reviewquestion}[1]{\textcolor{red!80!black}{[\textbf{Review question}:#1]}}
%\newcommand{\reviewquestion}[1]{\textcolor{red}{\begin{itemize}\item Question: #1\end{itemize}}}
%\newcommand{\reviewquestion}[1]{}

@@

@@

\makefrontmatter
\chapter{Introduction} % should give the background of your work. Important parts to cover:
% Give the context of your work, have a short introduction to the area.
While writing software, developers often make mistakes.
Catching these mistakes as early as possible can prevent these bugs from ever reaching production code.
%\CR{Possible papers to cite as evidence:
%  \url{https://dl.acm.org/doi/abs/10.1145/1985793.1985889}
%  \url{https://www.usenix.org/legacy/event/sec04/tech/full_papers/johnson/johnson_html/}
%  \url{https://ieeexplore.ieee.org/abstract/document/10190394}}
%\CR{These papers add static checkers on top of existing languages to find additional bugs; you can also
%  mention statically typed languages, but those generally don't require frequent changes to the type checker implementation, so they're less
%  compelling to motivate your work.}
Type checkers are one of the most common static checking methods, ensuring that the correct data types are used.
While some languages include type checkers in their compilers, work has also been done to add static type checkers to existing languages.
This requires writing large amounts of specialised code on a per language basis, but has shown to be effective at detecting bugs\cite{10.1145/1985793.1985889}\cite{10.5555/1251375.1251384}\cite{10190394}.
Furthermore, making changes to the typing rules of a language can require extensive rewrites in many different parts of the type checker's codebase.

% Define the problem you are solving (or trying to solve).
%\CR{Prior work has already looked into pluggable type checkers.
%  Why is translating them into JastAdd interesting?
%  (aside from theoretical questions, you can also use
%  pragmatic reasons, such as integration into our courses.)}
In this thesis we examine the feasibility of automating the writing of type checkers, based on the formal definition of the language's typing rules, using JastAdd's implementation of Reference Attribute Grammars (RAGs).
JastAdd is an interesting target because it has already been used to implement compilers of several languages, to which our typing rule implementations could be integrated.
JastAdd is also used extensively in courses at the department of computer science at LTH, where our project could serve as an educational tool, providing students an interactive way to define and test simple type systems.

Typing rules form a complex, often ambiguous, and infinite relation.
To be able to translate them into JastAdd code, we will need to define a limited subset of syntax-directed typing relations.
We identify a number of different kinds of typing rules, used in previous work in the field\ref{Pierce}, which share a common set of features for us to implement.
We then devise a strategy to translate these rules into snippets of iterative Java code, which can be evaluated by JastAdd.

%\CR{Make clear to the reader what the challenges are:
%      How are these two formalisms similar, how do they differ (at a high level)?
%      You can use a simple typing rule as an example if it helps (very optional).
%      Hint: look at determinism in evaluation (syntax-directedness), unification, but stick to intuition in this chapter; you can elaborate in Chapter 2 or later if you think it helps.}

\CR{Usually you then put a high-level summary of the rest of the thesis here (not mentioning chapters yet), something like ``we examine X and Y and introduce a strategy for Z.
  We evaluate the approach against Foo, and demonstrate Bar.
  Our case studies show Quux.
  Overall, this initial exploration shows \ldots''}

% Specify your contributions. What does this particular work/report bring to the research are or to the body of knowledge?
\section{Research questions}
To formulate our goals more concretely, we have set out to answer the following research questions:
%\CR{Give the RQs numbers so you can refer to them later.  Common form: RQ1, RQ2 etc.; some authors use RQ2.3 if they split RQ2 into multiple subquestions (this can happen later).}
\newcommand{\rqone}{RQ1: What kinds of typing rules can we translate using RAGs?}
\newcommand{\rqtwo}{RQ2: How can we implement the translations and what algorithms should we use for the output code?}
\newcommand{\rqthree}{RQ3: What are the challenges in translating typing rules into JastAdd-based RAGs?}
\begin{itemize}
  \item \rqone
    %\CR{``kinds'' of typing rules: right now the reader isn't aware that there even are different kinds (see my earlier comment on ``what the challenges are''-- sketching this out should suffice.)}
  \item \rqtwo
    %\CR{What does ``form'' mean, here?  How broad is ``implement''-- are you talking algorithms (which I think you are) or including technical challenges (which I think you have in the next RQ)?}
  \item \rqthree
    %\CR{I'd go with `challenges in translating' instead of `potential issues of translating', language-wise}
\end{itemize}

\section{Chapter outline}
%\CR{%paper $\to$ thesis; also,
 % this always comes at the very end of the introduction}
The rest of this thesis is outlined as follows.
% This chapter introduces the
Chapter 2 presents the necessary background information for the report.
Chapter 3 describes the design and implementation of our program, followed by demonstrations of its results in chapter 4.
Chapter 5 discusses the results in a more general sense, before the conclusion in chapter 6.

\chapter{Background}
%\section{Theory} % should contain the theoretical background needed to understand your work, if necessary.
\CR{General: the first paragraphs should explain what the chapter is about and how/why it is relevant, at a high level.  The reader should come out of it thinking ``I know why I want to read this''}
This chapter aims to provide all the background information required to understand the work presented in the rest of the report.
Section \ref{abstractsyntaxtrees} explains abstract syntax trees, a simplified tree-based representation of a program, suitable for performing static analysis on.
These are extensively utilised in almost every stage of our project, both in terms of our implementation and the input to the program.
%\CR{Consider saying just a little more about them here, for readers who have no idea what ASTs are and why they matter for
%    analysing programs.}
Section \ref{typetheory} details type systems,
%\CR{Too much detail in the following:}
by explaining each of the component pieces, followed by a description of their use within static type checking.
%\CR{You can use \texttt{\textbackslash label\{labelname\}} and \texttt{\textbackslash ref\{labelname\}} to reference (sub)secttions, that's usually clearer than ``next section''.}

\section{Abstract syntax trees}\label{abstractsyntaxtrees}
\CR{What's a syntax?}
An abstract syntax tree (AST) is a tree-based representation of the syntax of a program.
Each node in the tree represents a syntactic construct of the language, such as expressions, statements or declarations and the branches represent the structure.
It is abstract in that it neglects to include all the concrete details of the syntax, but rather focuses on the structure of the program.
Structural elements of the source code, such as parenthesis, commas and brackets, are no longer necessary, as the structure of the tree conveys this semantic information.

This makes it easier to write analysis based on the behaviour of a program, without concerning oneself with its exact formulation.
Several different formulations of a program with identical behaviour may be abstracted into identical ASTs.
The hierarchical structure of the tree lends itself especially well for recursive analyses, and eases the passing of information between connected structures by connecting them as parent and children nodes.
%\CR{what is ``this information'', concretely?}
A statement like \verb|if (true) then {1} else {0}| maybe instead be represented by an if-else node, with named branches pointing towards the condition and the two values.
%\CR{You could show an example, but at the very least I would like to read an argument for why the hierarchical structure of a tree is helpful.}
%\CR{You can find common examples for showing the value of ASTs over concrete syntax trees in the EDAN65 slides.}

\CR{Consider giving the syntax of a simple expression language in BNF here, since you're already discussing ``Type syntax'' later.}
\begin{figure}[h]
\begin{lstlisting}[]
Expr   -> Expr "+" Term | Term
Term   -> Term "*" Factor | Factor
Factor -> Int | "(" Expr ")"
\end{lstlisting}
  \caption{\reviewquestion{One like this? I'm unsure how to tie it in effectively.}}
  \label{typesyntaxspec}
\end{figure}

\section{Type theory}\label{typetheory}
In this section we will discuss the elements of type theory required to understand this thesis.
Firstly, we will discuss what types are and give some examples of their use.
Next, we will describe type systems, which form a relation between types and expressions.

\section{Values and expressions}
The main purpose of most computer software is to handle and manipulate data.
Each of these individual pieces of data is a value, which could be anything from the number 4 or the text `Hello!'.

Expressions are a broader term, indicating any syntactic structure which evaluates to a value.
This includes but is not limited to values, which evaluate to themselves, but also constructs of values, operators or function calls, etc.

\subsection{Data types}
%\CR{think about the connection between types and values here.  You may even want to talk about values in a separate subsection.}
Data types can be considered a categorisation of values which share common properties, such as supporting numerical operations or accessing a certain element in a list of characters.
%several different types of data in one program, such as numbers or text.
Most programming languages come with a number of built-in data types, from different kinds of numbers (integers, floating point) to strings or lists and arrays.

%\CR{if you discuss values, you can use the intuition of ``grouping values'' by certain properties (such as supported operations) to make your point clearer.}
Different parts of programs may have different constraints on the types of expressions, such as a print function requiring a string of text, or a plus function requiring some kind of number.
%\CR{Very good point about types of expressions, but again, this is likely easier to understand by linking expressions and values.
%  I strongly recommend using a definition of the form ``all values are expressions, but not all expressions are values''.}

\subsection{Type checking}
Verifying that these constraints are upheld can assist us in catching programming mistakes, preventing the wrong data types from being used in the wrong places, which may have lead to erroneous or unpredictable behaviour.
There are two main approaches to ensuring the constraints are upheld, dynamic and static type-checking.
As our project generates a static type checker, we will focus our explanations on that topic, but provide a quick overview of their differences here.
%\CR{Why would we even want to ensure that they are upheld?}

Dynamic checking is done performed runtime, with each type checked only when needed.
While this means no checks are performed unnecessarily, it makes it harder to verify the constraints over the whole program, as it won't be tested until each specific part is executed.

Static checking is performed on the code before it is run, usually as part of compilation, preventing many faulty programs from being run at all.
However, since it lacks knowledge of what happens at runtime, it's required to take a more conservative approach, flagging potential errors even in code which will never executed.
Static checkers analyse the structure of the source code, often in the form of an abstract syntax tree.

%\CR{Why would I want static checking?  Dynamic only checks when needed, isn't that more efficient?}

\CR{One mental image that helped me a lot to understand types is to see values  on one side (let's say on the left) and their types on the other (let's say the right).
  For each concrete (and possibly very complex) operator on the left-hand side (e.g., $(+)$), we then get a small number of typing rules on the right-hand side.
  This symmetry between ``concrete'' and ``abstract'' only works if you make values part of your cast of characters for this discussion.
  The value of this mental image to your approach then becomes immediate: since operators correspond to AST nodes and typing rules are abstractions over operator behaviour, typing rules will abstract over the program semantics that are captured by individual AST nodes.
}

\subsection{Type systems}
A type system typically consists of two parts, a type syntax, describing the available types of a language; and typing rules, denoting how types should be assigned to expressions within the language.
%\CR{Why do the types need a syntax?
%    (There are other approaches to types; for your own sanity, I recommend deferring to Pierce here and calling this ``typical approaches'')}

\subsubsection{Type syntax}
A type syntax, or set of types, is a definition of the valid types within a language.
\CR{Is ``type syntax'' what Pierce calls it?  Not sure I've heard that term before; I don't think it's wrong, personally, but please check in case he uses a different term.}
\reviewquestion{I believe Pierce simply refers to it as the "set of types". I seem to have got the term type syntax from EDAP15 \url{https://fileadmin.cs.lth.se/cs/Education/EDAP15/2022/web/skills/type-system.html}. In my opinion it works better to have a clear name for it, so I'm keeping it for now, unless you have further thoughts.}
It can consist of fundamental types, such as integers or booleans.
There may also be parameterised or composite types, for example lists may have a type parameter defining what values it can contain.
This enables differentiating lists that store integers from those that store strings, and ensuring that any item added to a list must be of the same type as the type parameter.
%\CR{A bit vague; consider giving an example.}
\begin{figure}[h]
\begin{lstlisting}[]
abstract Type;

Bool : Type;
Int  : Type;
\end{lstlisting}
  \caption{Type syntax consisting of two types, for boolean and integer values, written as a JastAdd AST specification (see Section \ref{astdef})}
  \label{typesyntaxspec}
\end{figure}

\subsubsection{Typing rules}
The typing relation, notated \lstinline{t : T}, is the link between the expressions of a language and the types of the type syntax.
%\CR{It is \emph{the} link, even.}
The left-hand side is an expression in the object language, typically written in its native syntax, the right-hand side a type from the aforementioned type syntax.

Typing rules are a manner of inference rules, which in their simplest form, consist of no more than a typing relation.
The typing rule \lstinline{T-True} seen in Figure \ref{typingruleexampletrue}, for example, simply assigns the type of \lstinline{Bool} to the expression \lstinline{true}.
Additionally, the typing rules in these examples have a name, given in parentheses next to the rule, though this is merely for reference.

\begin{figure}[h]
  \begin{equation}
    \tag{T-True}
    \text{true} : \text{Bool}
  \end{equation}
  \caption{Typing rule for a true value}
  \label{typingruleexampletrue}
\end{figure}

Typing rules may also have premises, a number of conditions that must hold for the typing rule to apply.
%\CR{\emph{that} must hold (``which'' relates a set to a superset: ``carrots, which are vegetables'', while ``that'' restricts a set to a subset: ``vegetables that are carrots'')}
The notational convention here is to put the premises above a bar, with the conclusion underneath it.

\begin{figure}[h]
  \begin{equation}
    \tag{T-If}
    \frac{
      t_1 : \text{Bool}\quad
      t_2 : \textsf{T}\quad
      t_3 : \textsf{T}
    }{
      \text{if } t_1 \text{ then } t_2 \text{ else } t_3 : \textsf{T}
    }
  \end{equation}
  \caption{Typing rule for an if-expression}
  \label{typingruleexampleif}
\end{figure}

\newcommand{\ty}[1]{\textcolor{blue!20!black}{\textsf{#1}}}
\newcommand{\tyvar}[1]{\textsf{#1}}
Figure \ref{typingruleexampleif} shows an example rule, typing an if-expression.
The bottom line, the conclusion, contains the keywords of the if-expression and three free variables $t_i$ which represent any term within the language.
Above the bar are three premises, placing constraints on the three variables, which must hold for the typing rule to apply.
The first constrains $t_1$ to the type bool, the second two constrain $t_2$ and $t_3$ to a type variable \tyvar{T}, which represents any single arbitrary type within the type syntax.
While \tyvar{T} may be any type, it must consistently be the same type within any single evaluation of the rule, i.e. the two terms must have the same type.
This \tyvar{T} also appears in the conclusion, as the resulting type of the if-expression, meaning the result of the entire expression will be that of the $t_2$ and $t_3$ terms.
%\CR{Consider typesetting T differently, e.g. as \tyvar{T}}
%\CR{\tyvar{T}, $t_1$, $t_2$, $t_3$ are \emph{free} variables here, logically speaking}.

\subsubsection{Evaluating typing rules}
While performing static type checking, the typing rules form the basis of our analysis.
By analysing the program, we attempt to match the syntactic form with the typing rules, to find the type of the expression.
If no matching typing rules can be found, a type error is produced.

The simplest program possible with the typing rules T-True and T-If, consists of the single value \lstinline{true}.
It can immediately be matched to the rule T-True, giving the type \lstinline{Bool}.

A slightly more complex case can be expressed as \lstinline{if true then true else true}.
The only rule that matches the structure, is T-If, with \lstinline{true} values taking the place of each of the $t_i$ variables.
To evaluate this rule, we are thus required to recursively evaluate the types each of our $t_i$ variables, to ensure they match the premises.
In this case, the type of $t_1$ evaluates to \lstinline{Bool}, matching the first premise of T-If.
The types of both $t_2$ and $t_3$, represented as the type variable \tyvar{T}, also evaluate to \lstinline{Bool}.
Since both the uses of this type variable within the premises are consistent, all three of the rule's premises have been fulfilled, and the entire expression can be typed as \tyvar{T}, which has been found to correspond to \lstinline{Bool}.

\begin{figure}[h]
  \begin{equation}
    \tag{T-Zero}
    \text{zero} : \text{Int}
  \end{equation}
  \caption{Typing rule for a zero value}
  \label{typingruleexamplezero}
\end{figure}

Adding a second value of a different type allows us to express incorrectly typed programs, for example the program \lstinline{if zero then zero else true}.
Recursively evaluating this according to T-If, we find that the premise concerning $t_1$ no longer upholds, as the type of it now evaluates to \lstinline{Int}.
Additionally, the type variable \tyvar{T} typed by both $t_2$ and $t_3$ now has contradictory use, being assigned both as the type \lstinline{Int} and \lstinline{Bool}.
As this program cannot be found to satisfy any set of typing rules in our type system, it is found to be invalid.

%\CR{Please explain the purpose of typing rules in static type checking.  Using just your two examples you would be able to show a well-typed program, and if you add one more rule (e.g. for integers), you can show a type error.  No need to build the full derivation tree, but it's important to mention recursion here.  This entire bit may be worth a separate subsubsection.}

\section{Reference Attribute Grammars}
This section introduces the subject of Reference Attribute Grammars (RAGs), first explaining the theory behind them, then introducing JastAdd, an implementation of RAGs within a Java meta-compilation system\cite{JastAdd}.

\subsection{Reference Attribute Grammars}
Reference Attribute Grammars are an extension to Attribute Grammars (AGs), a formalism used for specifying the static semantics of programming languages.
%\CR{Generally only for the static semantics (RAGs can also do dynamic semantics, if you allow circular attributes, but that's not worth mentioning here.)
%Please also connect ``semantics'' to ``type checking'' (via ``static semantics'')}
AGs provide a systematic way to define computations on the nodes of an abstract syntax tree, associating attributes with language constructs and are commonly used in the context of compiler construction, where they help define the translation of source code into executable code\cite{AGs}.

@@

\subsection{Non-Terminal Attributes}
\cite{higherorderattributegrammars}
\CR{Your example uses Non-Terminal Attributes (NTAs) \url{https://dl.acm.org/doi/pdf/10.1145/74818.74830}, which are
not a necessary feature for RAGs.  Please introduce them in a separate section: you can very directly connect them to the idea of ``type syntax''.}

\subsection{JastAdd}
JastAdd is an implementation of RAGs for Java, offering a flexible system for writing language analysis in an object-oriented fashion.\cite{JastAdd}
Our project uses it both for the implementation of our compiler and as a compilation target, as it outputs JastAdd code.
%\CR{It's worth telling the reader here why you are explaining some details about JastAdd.}

\subsubsection{AST specification}\label{astdef}
One of the key components of a JastAdd projects are the \lstinline{.ast} files, which declare the structure of the abstract syntax tree.
This specification is then used by JastAdd to generate Java classes representing the nodes of the AST.
These AST classes can have a variety of forms, supporting both abstract classes and subclasses, and can contain components such as Java types, tokens and child nodes, and these components may also be optional or list components.
For a detailed overview of the syntax, see the reference manual\cite{jastaddref}.

\subsubsection{RAG Modules}
\begin{figure}[h]
\begin{lstlisting}[language=jrag]
aspect TypingRules {
  syn Type Zero.type() {
    return new Int();
  }
}
\end{lstlisting}
  \caption{Snippet of a .jrag file, declaring a type attribute for the Zero AST node}
  \label{boolstrgen}
\end{figure}
The RAG modules written in \lstinline{.jrag} files, declare the attributes to be included in the classes generated by the AST specification.
The files use a syntax very similar to that of Java to declare attributes of a variety of different kinds, and the body of methods corresponds directly to Java code.

@@

%\subsection{Typing rule compiler}\label{sectionimplementation}

@@

\begin{figure}[]
\begin{lstlisting}[language=jrag]
aspect TypingRules {

@@

\section{Research questions}\label{sectionresearchquestions}
